# Maps {#sec-maps1}

::: {.callout-note appearance="simple"}
Today we will be focusing on the practice of data visualization.
:::

R has multiple packages that enable geospatial data visualization. Today, we create 
basic maps using the `sf` and `leaflet`.  `sf` stands for ![Simple Features](https://r-spatial.github.io/sf/articles/sf1.html). 

As a reminder, when we first use a package, we need to install it locally.  Let's start by installing both packages.
This only needs to be done once.

```{r}
#\ label: installation of leaflet and sf
#\ warning: false
#\ eval: false

install.packages('sf')
install.packages('leaflet')

```

Next, we load the packages to make sure they are available within the R environment.

```{r}
#\ label: installation of leaflet and sf
#\ warning: false


library(sf)
library(leaflet)

```

So far, this should be very familiar to what we did last week.

Now we'll import a test `sf`

```{r}
#\ label: NC dataset?
#\ warning: false

nc <- st_read(system.file("shape/nc.shp", package="sf"))

```

Let's

```{r}

library(tidyverse)


```